{\bf Practical Concurrency Testing} \\ \normalsize \vspace{1em} {\em or, How I Learned to Stop Worrying and Love the Exponential Explosion}

### 2017-04-03 proposal/410.tex
thesprop: split pintos and tm into their own chapters
garth revisions for thesprop

\chapter{Education}
\label{chap:410}

@@

In my MS thesis \cite{landslide}, I solicited students at CMU's Operating Systems Design and Implementation class
(henceforth, ``15-410'')
to volunteer at the end of the P3 project to annotate their kernels and try debugging them with Landslide.
However, the annotation burden undermined Landslide's purpose:
the only students willing to spend free time on manual instrumentation were biased to be those who were already doing well in the class,
and hence least likely to benefit from Landslide's debugging potential.
(Actually, even the best 15-410 students still have concurrency bugs,
but in principle, an educational tool must reach the more struggling students,
the so-called ``middle'' and/or ``bottom'' of the class.)
%
Requiring annotations hurts Landslide's case as a grading tool, as well:
TAs need to understand the kernel to begin with in order to annotate correctly,
and while achieving such understanding they may as well grade it by hand, as before.

Since then, I've extended Landslide to support testing P2 thread libraries (Section~\ref{sec:pebbles}) as well.
Because P2 mandates specific function names for the project's internal APIs
-- most importantly, for the concurrency primitives --
Landslide can automatically annotate arbitrary student implementations with no manual effort required of the user (whether student or TA).
The addition of Quicksand and Iterative Deepening (Chapter~\ref{chap:quicksand}) partly fulfills this purpose,
freeing student attention from the issue of which state spaces to test.
This chapter will detail my further techniques and evaluation which are specific to educational use.

@@

Every stateless model checker must make some assumptions about the tested programs' concurrency model \cite{chess}.
However, arbitrary programs may break conventional disciplines of concurrent programs, while still being bug-free.
For example, thread communication via ad-hoc {\tt yield} loops may appear as an infinite loop, livelock, or deadlock.
This is especially true of student code,
written by people who are just learning concurrent programming discipline for the first time,
and/or written under the time pressure of a project deadline.
To be an effective tool for struggling students,
a model checker must somehow coax adversarial programs to fit its concurrency model, to effectively test for real bugs,
rather than rejecting them outright on some stylistic or disciplinary grounds.

@@

@@

				Landslide marks the thread blocked (or ``disabled'', in model-checking parlance),

@@

		and also avoids the need to iteratively deepen the yield bound by fixing it as a heuristic constant.

@@

		When a deadlock arises under conditions where one or more threads are heuristically blocked,
		Landslide attempts to refute it as a false positive by artificially unblocking all heuristically-blocked threads%
		\footnote{If any threads are ICB-blocked, I prioritize waking those before trying to wake any yield-blocked threads.
		% Hmm... I don't remember exactly why.
		Waking all threads at once here can lead to unsoundness.}
		If the deadlock is true, each thread will immediately trigger the yield-blocking heuristic again,
		bringing the system back into deadlocked state.
		Landslide then repeats this process a heuristic constant number of times (128),
		allowing the program that many chances to make progress before proclaiming deadlock.
		(Note that this heuristic cannot miss true deadlocks as false negatives.)
		\footnote{Implemented in {\tt try\_avoid\_fp\_deadlock} in {\tt arbiter.c}.}
	\item {\bf Lock hand-off.}
		A common, though discouraged, idiom for implementing thread destruction involves one thread ``handing off'' ownership of a mutex to another.
		That thread will then release the lock with no corresponding acquire in its own execution.
		% In this case, some other thread communication enforces an ordering ... hmm,, pure vs limited hb?
		Although Landslide cannot easily recognize at what point the latter thread's accesses are protected by that lock for the sake of data-race analysis,
		potentially leading to false positive races,
		it must release the lock in its bookkeeping to avoid false {\em negatives} if any later access that should be protected by that lock isn't.
		When releasing a lock, if absent from the current thread's lockset, Landslide searches the locksets of all other existing threads, and releases it there.
		Landslide can instead optionally be configured to treat lock hand-off as an outright bug, as a matter of discipline.
		\footnote{Implemented in {\tt lockset\_remove} in {\tt lockset.c}.}

@@

However, some of the less common offenses are both more difficult to handle algorithmically, and also more worrying from a pedagogical point of view.
%After some collaboration with David Eckhardt (15-410 professor and member of this thesis committee),
For the following cases, we configured Landslide to abort, and warn the student that they must find a better solution before it could test their code.

@@

To analyze its effectiveness as a debugging tool for students,
I configured it to record a snapshot of the results of each use by the students:
which options were used (test program and time limit),
a snapshot of the student's code, and the result of the test
(completed, timed out, bug found, or ctrl-C'ed).
I manually interpreted the snapshots to determine:
\begin{itemize}
	\item How many unique bugs the student found (discounting multiple runs that produced ``the same'' bug);
	\item How many of those bugs were deterministic versus concurrency bugs (did Landslide find the bug on the first interleaving or did it need to test multiple);
	\item How many of each category of bug the students fixed (determined when a subsequent run of Landslide on the test case failed to find the bug again).
\end{itemize}

{\bf Results -- bugs found.}
Of 90 two-student groups in those 3 semesters, 47 tested their thread libraries with Landslide.
Table~\ref{tab:this-table-sucks-but-it's-the-best-i-got} shows
that Landslide found in total 44 deterministic bugs and 85 concurrency bugs.
It found at least one concurrency bug for 32 groups (68\%),
24 of whom (51\%) were able to fix at least one, verifying their update
with a succesful re-run of the same test.

The table's data presentation is somewhat confusing:
the dependent columns count the number of {\em groups}, categorized by how many bugs they found, not the raw number of {\em bugs},
which I instead indicate in the ``Total bugs'' row at the bottom.
So, the number of total bugs is equal to the sum of the products between each cell in that column and its corresponding number of bugs,
e.g., $44 = 27\times0 + 6\times1 + 9\times2 + 2\times3 + 1\times4 + 2\times5$.
Sorry about that; I'll try to find a better way to present this data in the thesis.

@@

Unfortunately, this turned out to be a negative result, as shown in Figure~\ref{fig:grade-distribution}.
The distribution of Landslide users' ultimate grades is indistinguishable from that of students for whom Landslide was unavailable.
While there was a 3\% increase in P2 grades between opt-out-ers and Landslide users, it's prone to selection bias:
perhaps those who opted out of Landslide were already the worst of the class who simply didn't have any free time for it.
The distribution of P3 grades is indistinguishable as well.
In the thesis, I will discuss possible factors contributing to this failure, and how they might be controlled for in future work,
although I am not planning to run any further experiments in the limited time I have.

@@

{\bf Future work.}
One evaluation question remains which I will answer in the thesis:
what is the accuracy of Landslide's automation heuristics?
In other words, were false-positive infinite loops or deadlocks ever reported
where Landslide failed to recognize an ad-hoc synchronization?
(I expect a near 100\% success rate,
having tuned the heuristics by hand on an (independent) set of P2 submissions already.)

\section{Landslide as a Grading Tool}
\label{sec:grading}

Beginning in 2017, I will recruit 15-410 Teaching Assistants (TAs) for my user study rather than students.
The modified user study will evaluate Landslide as a grading tool,
alternative to the ``Fritz'' stress-testing infrastructure which 15-410 currently employs,
and possibly substituting as well for the human process of reading student code looking for concurrency bugs by hand.
The intent is not for Landslide to ultimately replace 15-410's manual grading process,
but rather to demonstrate that comparably-rigorous grading could be deployed automatedly at other schools wishing to adopt the 15-410 curriculum.

I will pursue the following evaluation questions:

\begin{itemize}
	\item Provided appropriate test cases, can Landslide find the most common ``deep'' concurrency bugs in student submissions?
	\item How does Landslide's bug-finding compare to that of Fritz, the class's stress tester?
	\item How does Landslide's bug-finding compare to that of TAs grading by hand?
\end{itemize}

Finally, I will conclude this chapter in the thesis by discussing how future work could adapt Pebbles's interface to support fully-automatic model checking without compromising the educational value of P3.

>>>>

\chapter{Pebbles}
\label{chap:410}

@@

In my MS thesis \cite{landslide}, I solicited students at CMU's Operating Systems Design and Implementation class
(henceforth, ``15-410'')
to volunteer at the end of the P3 project to annotate their kernels and try debugging them with Landslide.
However, the annotation burden undermined Landslide's purpose:
the only students willing to spend free time on manual instrumentation were biased to be those who were already doing well in the class,
and hence least likely to benefit from Landslide's debugging potential.
(Actually, even the best 15-410 students still have concurrency bugs,
but in principle, an educational tool should reach the more struggling students,
the so-called ``middle'' and/or ``bottom'' of the class.)
%
Requiring annotations hurts Landslide's case as a grading tool, as well:
TAs need to understand the kernel to begin with in order to annotate correctly,
and while achieving such understanding they may as well grade it by hand, as before.

Since then, I've extended Landslide to support testing Project 2 (P2) thread libraries (Section~\ref{sec:pebbles}) as well.
Because P2 mandates specific function names for the project's internal APIs
-- most importantly, for the concurrency primitives --
Landslide can automatically annotate arbitrary student implementations with no manual effort required of the user (whether student or TA).
The addition of Quicksand and Iterative Deepening (Chapter~\ref{chap:quicksand}) partly fulfills this purpose,
freeing student attention from the issue of which state spaces to test.
This chapter will detail my further techniques and evaluation which are specific to educational use.

@@

Every stateless model checker must make some assumptions about the tested programs' concurrency model \cite{chess}.
However, arbitrary programs may break conventional disciplines of concurrent programs, while still being bug-free.
For example, thread communication via ad-hoc {\tt yield} loops may appear to automated tools as a possible
infinite loop, livelock, or deadlock.
This is especially true of student code,
written by people who are just learning concurrent programming discipline for the first time,
and/or written under the time pressure of a project deadline.
To be an effective tool for struggling students,
a model checker should somehow coax adversarial programs to fit its concurrency model, to effectively test for real bugs,
rather than rejecting them outright on some stylistic or disciplinary grounds.

@@

				The consequence of any this heuristic's inaccuracy is
				that an unusual wait-loop might be classified as an infinite loop bug anyway.

@@

				Landslide speculatively marks the thread blocked (or ``disabled'', in model-checking parlance),

@@

		and also avoids the need to iteratively deepen the yield bound by instead fixing it as a heuristic constant.

@@

		When a deadlock arises under conditions where one or more threads are heuristically blocked,
		Landslide attempts to refute it as a false positive by artificially unblocking all heuristically-blocked threads.
		\footnote{If any threads are ICB-blocked, I prioritize waking those before trying to wake any yield-blocked threads.
		% Hmm... I don't remember exactly why.
		Waking all threads at once here can lead to unsoundness.}
		Landslide then repeats this process a heuristic constant number of times (128),
		allowing the program that many chances to make progress before proclaiming deadlock.
		\footnote{Implemented in {\tt try\_avoid\_fp\_deadlock} in {\tt arbiter.c}.}
		Note that this heuristic cannot miss true deadlocks as false negatives:
		if the deadlock is true, each thread will immediately trigger the yield-blocking heuristic again,
		bringing the system back into deadlocked state as many times as necessary to exhaust the heuristic limit.
	\item {\bf Lock hand-off.}
		A common, though discouraged, idiom for implementing thread destruction involves one thread ``handing off'' ownership of a mutex to another.
		That thread will then release the lock with no corresponding acquire in its own execution.
		% In this case, some other thread communication enforces an ordering ... hmm,, pure vs limited hb?
		Although Landslide cannot easily recognize at what point the latter thread's accesses are protected by that lock for the sake of data-race analysis,
		potentially leading to false positive races,
		it must release the lock in its bookkeeping to avoid false {\em negatives} if any later access that should be protected by that lock isn't.
		When releasing a lock, if absent from the current thread's lockset, Landslide searches the locksets of all other existing threads, and releases it there.
		Landslide can instead optionally be configured to treat lock hand-off as a style warning or as an outright bug, as a matter of discipline.
		\footnote{Implemented in {\tt lockset\_remove} in {\tt lockset.c}.}

@@

However, some of the less common offenses are both more difficult to handle algorithmically, and also more worrying from a pedagogical point of view.
%After some collaboration with David Eckhardt (15-410 professor and member of this thesis committee),
For the following cases, I configured Landslide to abort, and warn the student that they must find a better solution before it could test their code (in other words, I promoted these patterns to be treated as bugs).

@@

To analyze its effectiveness as a debugging tool for students,
I configured Landslide to record a snapshot of the inputs and results of each use by the students:
which options were used (test program and time limit),
a snapshot of the student's code, and the result of the test
(completed, timed out, bug found, or ctrl-C'ed).
I manually interpreted the snapshots to determine:
\begin{itemize}
	\item How many unique bugs the student found (discounting multiple runs that produced ``the same'' bug);
	\item How many of those bugs were deterministic versus concurrency bugs (did Landslide find the bug on the first interleaving or did it need to test multiple);
	\item How many of each category of bug the students fixed (determined when a subsequent run of Landslide on the test case failed to find the bug again).
\end{itemize}

{\bf Results -- bugs found.}
Of 90 two-student groups in those 3 semesters, 47 tested their thread libraries with Landslide.
Table~\ref{tab:this-table-sucks-but-it's-the-best-i-got} shows
that Landslide found in total 44 deterministic bugs and 85 concurrency bugs.
It found at least one concurrency bug for 32 groups (68\%),
24 of whom (51\%) were able to fix at least one bug, verifying their update
with a succesful re-run of the same test.
\footnote{The table's data presentation is somewhat confusing:
the dependent columns count the number of {\em groups}, categorized by how many bugs they found, not the raw number of {\em bugs},
instead indicated in the ``Total bugs'' row at the bottom.
So, the number of total bugs is the sum of the products between each cell in that column and its corresponding number of bugs,
e.g., $44 = 27\times0 + 6\times1 + 9\times2 + 2\times3 + 1\times4 + 2\times5$.
%Sorry about that;
I'll try to find a better way to present this data in the thesis.}
I view this as a positive result -- among other statistics, Landslide was able to help the majority of students debug, among those who found concurrency bugs at all -- although it is difficult to control for the amount of time students spent with Landslide that would otherwise have been spent on old-fashioned stress testing.

@@

Unfortunately, this turned out to be a negative result, as shown in Figure~\ref{fig:grade-distribution}.
The distribution of Landslide users' ultimate grades is indistinguishable from that of students for whom Landslide was unavailable.
While there was a 3\% increase in P2 grades between opt-out-ers and Landslide users, it's prone to selection bias:
perhaps those who opted out of Landslide were already the worst of the class who simply didn't have any free time for it.
The distribution of P3 grades is indistinguishable as well.
In the thesis, I will discuss possible factors contributing to this failure, and how they might be controlled for in future work.
%although I am not planning to run any further experiments in the limited time I have.

@@

{\bf Future work.}
One evaluation question remains which I will answer in the thesis:
what is the accuracy of Landslide's automation heuristics?
In other words, were false-positive infinite loops or deadlocks ever reported
where Landslide failed to recognize an ad-hoc synchronization?
(I expect a near 100\% success rate,
having tuned the heuristics by hand on an (independent) set of P2 submissions already.)
%
%\section{Landslide as a Grading Tool}
%\label{sec:grading}
%
%Beginning in 2017, I will recruit 15-410 Teaching Assistants (TAs) for my user study rather than students.
%The modified user study will evaluate Landslide as a grading tool,
%alternative to the ``Fritz'' stress-testing infrastructure which 15-410 currently employs,
%and possibly substituting as well for the human process of reading student code looking for concurrency bugs by hand.
%The intent is not for Landslide to ultimately replace 15-410's manual grading process,
%but rather to demonstrate that comparably-rigorous grading could be deployed automatedly at other schools wishing to adopt the 15-410 curriculum.
%
%I will pursue the following evaluation questions:
%
%\begin{itemize}
%	\item Provided appropriate test cases, can Landslide find the most common ``deep'' concurrency bugs in student submissions?
%	\item How does Landslide's bug-finding compare to that of Fritz, the class's stress tester?
%	\item How does Landslide's bug-finding compare to that of TAs grading by hand?
%\end{itemize}
%
Finally, I will conclude this chapter in the thesis by discussing how future work could adapt Pebbles's interface to support fully-automatic model checking without compromising the educational value of P3.

### 2017-04-03 proposal/pintos.tex
garth revisions for thesprop

The existing Pintos support includes
a new set of annotations (e.g., {\tt mutex\_lock} in Pebbles is now called {\tt sema\_down};
handling several quirks of the basecode's scheduling behaviour; %(ask me if you really care to know);
extending the heap tracker to handle Pintos's page allocator {\tt palloc} as well as the kernel {\tt malloc},
as well as the fact that kernel memory is not direct-mapped like in Pebbles;
new thread-liveness code to detect when a test case terminates;
skipping some busy-wait loops in the boot sequence used for device communication,
an extra fall-back case in Landslide's timer injection algorithm, %the workings of which I don't even remember anymore,
and extending the lock-set and vector-clock analyses to include disabling interrupts.
%(which was actually pretty tricky when using vector clocks for pure-HB).
All these features apply to behaviours included with the stock base-code, so should apply generally to all but the most adventurous student implementations.

@@

Most practically, I will need to free Landslide from its dependence on Simics,
which requires paid licenses for its use. %beyond CMU's walls.
Other candidate emulation (simulation or virtualization) platforms for Landslide include Bochs, QEMU, VMWare, and Xen.
I am presently studying which of these will most suit Landslide's needs,
but so far Bochs seems like the best bet,
which the Berkeley, Stanford, and U. of Chicago classes already use as their simulator.

@@

Having only two semesters to test Landslide with Pintos students,
I'll be unable to gather as comprehensive a dataset as I have with 15-410 at CMU.
This will serve more as a proof-of-concept, demonstrating that MC can realistically replace stress testing in operating systems courses worldwide.
Nevertheless, I intend also to analyze what little data I can gather, comparing the incidences and types of bugs found between Pebbles and Pintos populations.
This will perhaps shed light on the advantages and/or shortcomings of either project,
leading to recommendations for improving either project's educational value (independently of using MC).

>>>>

So far, just enough Pintos support exists to have run Chapter \ref{chap:quicksand}'s experiments. This includes:
%The existing Pintos support includes
\begin{itemize}
	\item a new set of annotations (e.g., {\tt mutex\_lock} in Pebbles is now called {\tt sema\_down};
	\item handling several quirks of the basecode's scheduling behaviour; %(ask me if you really care to know);
	\item extending the heap tracker to handle Pintos's page allocator {\tt palloc} as well as the kernel {\tt malloc},
		as well as the fact that kernel memory is not direct-mapped like in Pebbles;
	\item new thread-liveness code to detect when a test case terminates;
	\item skipping some busy-wait loops in the boot sequence used for device communication;
	\item an extra fall-back case in Landslide's timer injection algorithm; %the workings of which I don't even remember anymore,
	\item and extending the lock-set and vector-clock analyses to include disabling interrupts.
\end{itemize}
%(which was actually pretty tricky when using vector clocks for pure-HB).
All these features apply to behaviours included with the stock base-code, so should apply generally to all but the most adventurous student implementations.

@@

Most practically, I will need to free Landslide from its dependence on Simics,
which requires paid licenses for its use. %beyond CMU's walls.
Other candidate emulation (simulation or virtualization) platforms for Landslide include Bochs, QEMU, VMWare, and Xen.
%I am presently studying which of these will most suit Landslide's needs,
%but so far Bochs seems like the best bet,
I have begun work porting Landslide to Bochs, which is open-source and provides both instruction and memory tracing, and
which the Berkeley, Stanford, and U. of Chicago classes already use as their simulator.

@@

%Having only two semesters to test Landslide with Pintos students,
%I'll be unable to gather as comprehensive a dataset as I have with 15-410 at CMU.
%This will serve more as a proof-of-concept, demonstrating that MC can realistically replace stress testing in operating systems courses worldwide.
%Nevertheless, I intend also to analyze what little data I can gather, comparing the incidences and types of bugs found between Pebbles and Pintos populations.
%This will perhaps shed light on the advantages and/or shortcomings of either project,
%leading to recommendations for improving either project's educational value (independently of using MC).

I aim to run a similar user study with Pathos students as I've done so far with Pebbles.
After finding one or more Pintos-using classes at other schools to collaborate with,
I will distribute Landslide to the students during the 'threads' and 'userprog' projects,
have them run a similar suite of tests to those in Chapter \ref{chap:quicksand}'s experiments,
and collect and analyze the results.
I plan to analyze the overall incidences and types of bugs found between Pebbles and Pintos populations,
which will perhaps shed light on the advantages and/or shortcomings of either project,
and lead to recommendations for improving either project's educational value (whether by incorporating MC or otherwise).

### 2017-04-03 proposal/related.tex
write brief eval plan for tm
garth revisions for thesprop

To my knowledge, this is the first study of model checking in an educational setting, although teaching concurrency is not itself an unstudied problem.
\cite{towards-a-framework} presents a promising framework for testing concurrent programs,
which can incorporate model checking as well as static analysis, resource exhaustion, data-race analysis, and coverage analysis.
However, it lacks an evaluation, and makes mention of its educational value only in its future work remarks.
%
\cite{how-studence} presents a survey of how students think about concurrent program behaviour and debugging,
while \cite{learning-concurrency} investigates how students form cognitive patterns about concurrent programming that could either aid or stunt their reasoning.
Both of these studies could help optimize Landslide's bug reports for clarity and student enlightenment.

@@

Replay analysis extends single-pass data-race analysis to apply model checking (in limited quantities) to classify data-race candidates by their impact on program behaviour.
It was first introduced in \cite{recordreplaydrs}, which compares the program states immediately after the access pair for differences, preferring still to err on the side of false positives.
RaceFuzzer \cite{racefuzzer} avoids false positives by requiring an actual failure be exhibited, as we do,
although it uses random schedule fuzzing rather than model checking.
%
Portend \cite{portend} is closest in spirit to Quicksand:
it tests alternate executions based on single-pass data-race candidates to classify them in a taxonomy of likely severity.
It uses symbolic execution to test input nondeterminism as well as schedule nondeterminism,
although its verification properties are not as strong as Quicksand's.

@@

Transactional memory (TM), first introduced in \cite{transactional-memory},
has received renewed attention in recent years with the announcement of Intel's Haswell architecture \cite{htm-haswell},
which promises to support hardware transactions using new x86 instructions.
Because hardware transactions can fail for any reason, not just on memory conflicts,
software TM remains relevant,
and recent works such as \cite{hybrid-htm-stm} and \cite{stm-relaxed-memory} enhance it to be nested with hardware transactions, or to be used on weak memory architectures, respectively;
which in turn produce ever more complicated concurrency models for people like me to tackle.
%
Testing approaches for transactional programs are sparsely represented in the literature so far.
%The Weakest Reasonable Condition
Several related works \cite{tm-correctness,tm-completeness,specifying-verifying-tm} are building up to formal proofs of the correctness of TM {\em implementations},
but not that of client programs which use TM.
Getting closer, McRT STM \cite{mc-tm-with-spin} uses SPIN \cite{spin} to model check an STM implementation
up to 2 threads running 1 transaction each with up to 3 memory accesses.
This kind of verification is an important stepping stone for trusting the results Landslide will provide.
% TODO: make this point once you've added discussion of mutex_test to chap:quicksand.
% This kind of verification is analogous to the {\tt mutex\_test}
%
{\em Learning from Mistakes} \cite{learning-from-mistakes} studies the characteristics of many types of concurrency bugs;
it found that TM could potentially fix some, but not all, of the studied bugs,
although in some cases it must be combined with other concurrency primitives,
which indicates a major need for verifying transactional code with model checking.

>>>>

To my knowledge, this is the first study of model checking in an educational setting, although teaching concurrency is not itself an unstudied problem.
Eytani et al.~\cite{towards-a-framework} present a promising framework for testing concurrent programs,
which can incorporate model checking as well as static analysis, resource exhaustion, data-race analysis, and coverage analysis.
However, it lacks an evaluation, and makes mention of its educational value only in its future work remarks.
%
L\"{o}nnberg et al.~\cite{how-studence} present a survey of how students think about concurrent program behaviour and debugging,
while Kolikant \cite{learning-concurrency} investigates how students form cognitive patterns about concurrent programming that could either aid or stunt their reasoning.
Both of these studies could help optimize Landslide's bug reports for clarity and student enlightenment.

@@

Replay analysis extends single-pass data-race analysis to apply model checking (in limited quantities) to classify data-race candidates by their impact on program behaviour.
It was first introduced by Narayanasamy et al.~\cite{recordreplaydrs}, which compares the program states immediately after the access pair for differences, preferring still to err on the side of false positives.
RaceFuzzer \cite{racefuzzer} avoids false positives by requiring an actual failure be exhibited, as we do,
although it uses random schedule fuzzing rather than model checking.
%
Portend \cite{portend} is closest in spirit to Quicksand:
it tests alternate executions based on single-pass data-race candidates to classify them in a taxonomy of likely severity.
It uses symbolic execution to test input nondeterminism as well as schedule nondeterminism,
although its verification properties are not as strong as Quicksand's.

@@

\section{Transactional Memory}
\label{sec:related-tm}

Transactional memory (TM), first introduced in 1993 \cite{transactional-memory},
has received renewed attention in recent years with the announcement of Intel's Haswell architecture \cite{htm-haswell},
which supports hardware transactions using new x86 instructions.
Because hardware transactions can fail for any reason, not just on memory conflicts,
software TM remains relevant,
and recent works \cite{hybrid-htm-stm,stm-relaxed-memory} enhance it to be nested with hardware transactions, or to be used on weak memory architectures, respectively;
which in turn produce ever more complicated concurrency models for people like me to tackle.
%
Testing approaches for transactional programs are sparsely represented in the literature so far.
%The Weakest Reasonable Condition
Several related works \cite{tm-correctness,tm-completeness,specifying-verifying-tm} are building up to formal proofs of the correctness of TM {\em implementations},
but not that of client programs which use TM.
SI-TM \cite{si-tm} introduces techniques for reducing HTM's abort rates, but without fully eliminating aborts, MCs must still consider them possible anywhere.
Getting closer, McRT STM \cite{mc-tm-with-spin} uses SPIN \cite{spin} to model check an STM implementation
up to 2 threads running 1 transaction each with up to 3 memory accesses.
This kind of verification is an important stepping stone for trusting the results Landslide will provide.
% TODO: make this point once you've added discussion of mutex_test to chap:quicksand.
% This kind of verification is analogous to the {\tt mutex\_test}
%
{\em Learning from Mistakes} \cite{learning-from-mistakes} studies the characteristics of many types of concurrency bugs;
it found that TM could potentially fix some, but not all, of the studied bugs,
although in some cases it must be combined with other concurrency primitives,
which indicates a major need for verifying transactional code with model checking.

### 2017-04-03 proposal/thesprop.tex
garth revisions for thesprop

Stateless model checking is a powerful technique for testing concurrent programs,
capable in theory of rooting out any bug or providing total verification on any program,
but suffers several problems which relegate that theory to fantasy land \cite{vargomax}.
Chief among those problems is exponential explosion of state spaces,
making it difficult to decide in advance which combinations of preemption points can be productively tested within a given time limit.
Another major problem is the manual annotation effort required to test certain types of concurrent programs,
which is especially relevant for operating systems students who implement their own kernels.
In this document I have proposed a thesis which will solve both problems.
I leave you now with a proposed timeline for bringing each project to fruition.

>>>>

Stateless model checking is a powerful technique for testing concurrent programs,
capable in theory of rooting out any nondeterministic bug or providing total verification on any program,
but suffers several problems which relegate that theory to fantasy land \cite{vargomax}.
Chief among those problems is exponential explosion of state spaces,
making it difficult to decide in advance which combinations of preemption points can be productively tested within a given time limit.
Another major problem is the manual annotation effort required to test certain types of concurrent programs,
which is especially relevant for operating systems students who implement their own kernels.
In this document I have proposed a thesis which will solve both problems.
I leave you now with a proposed timeline for bringing each project to fruition.
